One-arm spirals in self-gravitating dead zones in protoplanetary discs

### 2014-10-22 paper/model.tex
minor updates

The bump function
$B(R)$ represents a surface density boost between
$R\in[R_{d1},R_{d2}]$ by a factor $\epsilon^{-1}$ ($\epsilon < 1$),
and $\delta R_d$ is the transition width between the bump and the
smooth disc. We choose 
\begin{align}
  &B(R) = f_1(R)\times f_2(R),\\
  &f_1(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 +
    \tanh\left(\frac{R-R_{d1}}{\Delta_1}\right)\right]  + \epsilon,\\
  &f_2(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 -
    \tanh\left(\frac{R-R_{d2}}{\Delta_2}\right)\right]  + \epsilon,
\end{align}
where $\Delta_1 = \delta R_d \times H(R_{d1})$ and similarly for
$R_{d2}$. The bump function mimicks mass built-up in a dead zone.

@@

\section{Fiducial simulations}
Here we describe one numerical simulation from each code to give an
overall picture of disc evolution. In the next section, we analyse 2D
simulations in more detail and explore parameter space.

In these simulations we subject the disc to random perturbations in
cylindrical velocity,
\begin{align}
  \delta v_R = \frac{\delta}{M}c_s(R)\sum_{m=1}^M\cos{m\phi},
\end{align}
where $\delta\in[-10^{-3},10^{-3}]$ is chosen randomly but does not
depend on $\phi$, and we set $M=10$.

\subsection{FARGO 2D}
We consider a disc size $[R_\mathrm{min}, R_\mathrm{max}] =
[0.4,10]R_0$ and resolution $N_R\times N_\phi = 1024\times 2048$ or
about $16$ grids per $H$. We adopt $\epsilon_g=10^{-4}H$ for the 
self-gravity softening length\footnote{In 2D self-gravity, $\epsilon_g$ also
approximates for the vertical disc thickness. A more appropriate
value would be $\sim H$ \citep{muller12}. However, because
$\epsilon_g\propto R$ is needed in FARGO, the Poisson kernel
(Eq. \ref{2d_grav}) is no longer symmetric. We choose a
small  
$\epsilon_g$ in favour of angular momentum conservation, keeping in
mind that the strength of self-gravity will be over-estimated.}.

\subsection{ZEUS-MP 3D}

\subsection{PLUTO 3D}

>>>>

The bump function
$B(R)$ represents a surface density boost between
$R\in[R_{d1},R_{d2}]$ by a factor $\epsilon^{-1}>1$,
and $\delta R_d$ is the transition width between the bump and the
smooth disc. We choose 
\begin{align}
  &B(R) = f_1(R)\times f_2(R),\\
  &f_1(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 +
    \tanh\left(\frac{R-R_{d1}}{\Delta_1}\right)\right]  + \epsilon,\\
  &f_2(R) = \frac{1}{2}\left(1 - \epsilon\right)\left[1 -
    \tanh\left(\frac{R-R_{d2}}{\Delta_2}\right)\right]  + \epsilon,
\end{align}
where $\Delta_1 = \delta R_d \times H(R_{d1})$ and similarly for
$R_{d2}$. The bump function mimicks mass built-up in a dead zone.

@@

### 2014-10-22 paper/paper.tex
minor updates

\maketitle
\begin{abstract}

\end{abstract}

@@

\input{model}
%\input{results}
%\input{discussion}
%\input{summary}

>>>>

\maketitle
\begin{abstract}
  Theoretical models of circumstellar or protoplanetary discs indicate
  the possible existence of `dead zones' --- regions in the disc
  where mass accretion is greatly reduced because of inefficient
  magneto-hydrodynamic turbulence that would otherwise enable 
  accretion. It has been suggested that mass built-up in a dead zone
  may lead to gravitational instability. However, the type of
  gravitational instability relevant in this context has not been
  clarified. In this work, we use numerical simulations to explore
  gravitational instability in astrophysical discs in which only a 
  sub-portion of it is self-gravitating, taken as a toy model for
  protoplanetary disc containing a dead zone. We conduct numerical hydrodynamic
  experiments in 2D using the FARGO code and in 3D with the ZEUS-MP
  and PLUTO codes. In all cases we observe the development of a spiral
  mode with azimuthal wavenumber $m=1$ inside the dead
  zone. Surprisingly, the co-rotation radius of the spiral is well
  outside the dead zone, resulting in a nearly stationary one-arm spiral
  pattern in the dead zone. Our simulations demonstrate the potential
  importance global gravitational interaction between a
  self-gravitating dead zone and the exterior, non-self-gravitating
  parts of the disc.  
\end{abstract}

@@

\input{model}
\input{results}
%\input{discussion}
%\input{summary}
